---
title: Main code for the = Emergent life-history strategies in a general condition-dependent
  ectotherm growth model with energetic cost of reproduction
author: "Asta Audzijonyte & Shane A. Richards"
date: "23 October 2017"
output:
  html_document: default
---

## Growth model

### Useful functions

```{r message=FALSE, warning=FALSE}
rm(list = ls()) # clear memory
library(tidyverse)
library(tibble)
library(ggplot2)


# Reserve:Struture ratio 
RSRatio <- function(age) {
  tmp <- rs1*(age-a_bar)
  tmp <- max(tmp, -20) # bound below
  tmp <- min(tmp,  20) # bound above

  return(rs_min + (rs_max - rs_min)*exp(tmp) / (1.0 + exp(tmp)))
} 

# Predator length [m]
predatorLength <- function(S) {
  return((S/l_const)^(1/3.0)) # assumes invariant growth
} 

# Mortality rate [d-1]
mortProb <- function(l) {
  m_rate <- m_min + (m_max-m_min)*exp(-m1*l) # instantaneous mortality rate
  return(1.0 - exp(-m_rate)) # probability die per day
} 

# intake rate [d d-1]
grossIntake <- function(S) {
  return(g0*S^g1)
}

# DEB maintenance version [g d-1 g-1]
maintenance <- function(S, R) {
  return(ms*S + mr*w*R)
} 

# Energetic reproductive cost
reproCost <- function(S) {
  return(ra*S^rb)
}
```

### Model parameters for baseline zero fishing scenario

```{r message=FALSE, warning=FALSE}

# Model parameters for the baseline zero fishing scenario (First scenario in Table S2)

max_age_years <- 20 # 20        # years of simulation
max_age_days  <- 365*max_age_years # (d)

rs_max  <- 1.3           # maximum RS ratio
rs_min  <- 0.0           # minimum RS ratio

## length-weight conversion: 
l_const <- 1250/(0.60^3) # (g m-1) num = weight (g), denom = length (m)
#length-weight conversion uses weight of S only and assumes that 1250g of S weight (ca 3000g of total) corresponds to 60cm long fish

g0      <- 0.1           # intake rate constant: intake when structural weight is 1 g (g d-1), includes assimilation efficiency 
g1      <- 0.67          # power to uptake rate with S weight

ms      <- 0.003         # maintenance cost of structural mass (g d-1 g-1)
mr      <- 0.0003        # maintenance cost of reversible mass (g d-1 g-1)  

s_eff   <- 0.33          # conversion efficiency of assimilated intake to structure
r_eff   <- 0.9           # conversion efficiency of assimilated intake to reversible pool

# Reproductive cost function 
ra      <- 6             # reprod cost for 1 g of struct weight: (g g-1)
rb      <- 0.6          # reproductive cost power w.r.t. structural weight

#Mortality parameters 
m_min   <- 0.1/365      # background mortality rate (d-1) 

m1      <- 8.0          # steepness of the length-dependent mortality rate 
m_max   <- 4/365        # maximum length-dependent mortality rate (d-1) [when zero length]

s1      <- 7.0          # steepness of the condition realted mortality rate [low value = higher stochasticity]
s_max   <- 4/365        # maximum condition related rate (d-1) [when R = 0]

#Fishing mortality parameters 
Fm = 0/365            #Instantaneous fishing mortality of fully recruited fish (day-1)
Fmid = 0.3              #length (in meters) of the 50% fishing selectivity
Fk = 20                 # steepness of the logistic fishing function 

#Three parameters for optimal life-history strategy must be optimised for different mortality values in another code. These values are for the baseline zero fishing scenario with above parameters

rs1     <- 0.002   # age-dependence (d-1) [provided by grid]
a_bar   <- 350          # age at mid-point (d) [provided by grid]
w       <- 0.6        # fraction of reserve which is reproductive (fixed)


```

### Plots to explore age specific R/S allocation, expected survival and reproduction cost

```{r fig.height = 3, fig.width=5}

# bounds for plotting
l_min   <- 0.0           # minimum length for plotting (m)
l_max   <- 0.45          # maximum length for plotting (m)
w_min   <- 1             # minimum mortality rate for plotting (d-1)
w_max   <- 10000          # maximum mortality rate for plotting (d-1)
  
# prepare data frame to display RS ratio
vec_age <- seq(from = 0, to = max_age_days, by = 1)
df_RS <- tibble(Age = vec_age, RSratio = sapply(vec_age, FUN = RSRatio))
ggplot(df_RS, aes(x = Age, y = RSratio)) + geom_line() + 
  ylim(0,rs_max) + xlab("Age (days)") + ylab("Ratio (Reserve:Structure)") +
  labs(title = "Strategy: desired ratio of structural to reserve mass") +
  theme_bw()

# prepare data frame to display survival curve
vec_length <- seq(from = l_min, to = l_max, length.out = 101)
df_Length <- tibble(Length = vec_length, Prob = mortProb(vec_length))
df_Length <- mutate(df_Length, Survive_Y = 1.0/(Prob*365.0))
ggplot(df_Length, aes(x = Length, y = Survive_Y)) + geom_line() + 
  labs(title = "Starvation independent survivorship") +
  xlab("Length (m)") + ylab("Expected years survive") + theme_bw()

# prepare data frame to display reproductive costs
vec_weight <- seq(from = w_min, to = w_max, length.out = 101)
df_weight <- tibble(Weight = vec_weight, Cost = reproCost(vec_weight))
ggplot(df_weight, aes(x = Weight, y = Cost)) + geom_line() + 
  geom_abline(intercept = 0, slope = 1, color = "grey") +
  labs(title = "Cost of reproduction (mass not converted to spawn)") +
  xlab("Structural weight (g)") + ylab("Reproductive cost (g)") + theme_bw()



```


### Main calculations

```{r}
# useful age-dependent values
Res               <- array(data=0,c(max_age_years,365)) # reserve mass (g)
Str               <- array(data=0,c(max_age_years,365)) # structural mass (g)
dayIntake         <- array(data=0,c(max_age_years,365))
dayMaintenance    <- array(data=0,c(max_age_years,365))
dayNetIntake      <- array(data=0,c(max_age_years,365))
dayLambda         <- array(data=0,c(max_age_years,365))
dayPredatorLength <- array(data=0,c(max_age_years,365))
daySurvival       <- array(data=0,c(max_age_years,365))
yearSpawn         <- rep(0, max_age_years)
yearRepCost       <- rep(0, max_age_years)
yearFitness       <- rep(0, max_age_years)
natmortality      <- array(data=0,c(max_age_years,365))
fishmortality     <- array(data=0,c(max_age_years,365))
dayGrowth         <- array(data=0,c(max_age_years,365))
relGrowth         <- array(data=0,c(max_age_years,365))
strGrowth         <- array(data=0,c(max_age_years,365))

# set initial weight
#S0       <- 1             # initial structural weight on day 0 (g)
#R0       <- S0*RSRatio(0) # enforce correct initial reserve-structural ratio

S0 <- 1/(1+RSRatio(0))
R0 <- RSRatio(0) *S0

Res[1,1] <- R0
Str[1,1] <- S0

daySurvival[1,1] <- 1.0 # initially all individuals are alive

  
# perform the simulation
for (yr in 1:max_age_years) {
  for (day in 1:364) {
    Rstart <- Res[yr,day] # starting mass (reserve)
    Sstart <- Str[yr,day] # starting mass (structure)
    
    dayPredatorLength[yr,day] <- (Sstart/l_const)^(1/3.0) # calc length using str
    fish_mort <- Fm / (1+exp(-Fk*(dayPredatorLength[yr,day]-Fmid)))
    mort_rate <- 
      m_min + (m_max-m_min)*exp(-m1*dayPredatorLength[yr,day]) + # non-starve
      s_max*exp(-s1*Rstart/Sstart)   +                            # starve
      fish_mort                                                   #fishing mortality 
    fishmortality[yr,day+1] <- fish_mort   
    mort_prob <- 1.0 - exp(-mort_rate) # probability die this day
    daySurvival[yr,day+1] <- (1-mort_prob)*daySurvival[yr,day] # prob alive
    natmortality[yr,day+1] <-
      m_min + (m_max-m_min)*exp(-m1*dayPredatorLength[yr,day]) + # non-starve
      s_max*exp(-s1*Rstart/Sstart)

  
    intake      <- g0*Sstart^g1                 # (g d-1)
    respiration <- ms*Sstart + mr*Rstart        # (g d-1)
    net_intake  <- intake - respiration         # (g d-1)

    dayIntake[yr, day]      <- intake
    dayMaintenance[yr, day] <- respiration
    dayNetIntake[yr, day]   <- net_intake
    age                     <- 365*(yr-1) + day # age of animal (days)
      
    # add bounds to prevent numerical issues when calculating lambda
    tmp <- rs1*(age-a_bar)
    tmp <- max(tmp, -20) # bound below
    tmp <- min(tmp,  20) # bound above
    dayLambda[yr, day] <- rs_min + (rs_max - rs_min)*exp(tmp) /
      (1.0 + exp(tmp)) # RS ratio = startegy
    
    if (net_intake >= 0) {
    
        dR <- r_eff*net_intake # maximum R allocation
        dS <- s_eff*net_intake # maximum S allocation
  # use Lambdamax istead of dayLambda and set w=1  
        if (dayLambda[yr, day]*Sstart > Rstart) { # need to bump up reserves
          r_take <- min(dayLambda[yr, day]*Sstart - Rstart, dR) 
          Rstart <- Rstart + r_take
          net_intake <- net_intake - r_take/r_eff
        } else { # need to bump up structure
          s_take <- min(Rstart/dayLambda[yr, day] - Sstart, dS) 
          Sstart <- Sstart + s_take
          net_intake <- net_intake - s_take/s_eff
        }
     # partition remaining mass to keep desired ratio
    Res[yr,day+1] <- Rstart + dayLambda[yr, day]*r_eff*s_eff*net_intake / (r_eff + dayLambda[yr, day]*s_eff) 
    Str[yr,day+1] <- Sstart + r_eff*s_eff*net_intake / (r_eff + dayLambda[yr, day]*s_eff)       
    
    } else {
    
     dR <- (1/r_eff)*net_intake # this is what should be taken from R given the conversion inefficiencies 
     newR = (Rstart + dR)
       if (newR < 0) {
          newR = 0
        }
     
     Res[yr,day+1] <- newR
     Str[yr,day+1] <- Sstart

  }  

    Growth = (Res[yr,day+1]+Str[yr,day+1]) - (Res[yr,day]+Str[yr,day]) #change in weight over the day
    percGrowth = (Growth/(Res[yr,day]+Str[yr,day]))*100
    sGrowth = Growth/Str[yr,day]*100
    dayGrowth[yr,day] = Growth
    relGrowth[yr,day] = percGrowth
    strGrowth[yr,day] = sGrowth
    
}

  dayPredatorLength[yr,365] <- dayPredatorLength[yr,364]
  dayGrowth[yr,365] <- dayGrowth[yr,364]
  relGrowth[yr,365] <- relGrowth[yr,364]
  strGrowth[yr,365] <- strGrowth[yr,364]

  # perform spawning
  repro_cost <- ra*Str[yr,365]^rb # fixed cost of reproduction
  spawn_mass <- max(0, w*Res[yr,365] - repro_cost) # spawning mass after cost
  
  if (spawn_mass > 0) { # enough to spawn?
    Res[yr,365]     <- Res[yr,365] - spawn_mass - repro_cost
    yearSpawn[yr]   <- spawn_mass
    yearRepCost[yr] <- repro_cost
    yearFitness[yr] <- spawn_mass*daySurvival[yr,365]
  } 
  
  if (yr < max_age_years) {
    Str[yr+1,1]         <- Str[yr,365]
    Res[yr+1,1]         <- Res[yr,365]
    daySurvival[yr+1,1] <- daySurvival[yr,365]
    natmortality[yr+1,1]  <- natmortality[yr,365]
    fishmortality[yr+1,1]  <- fishmortality[yr,365]
  }
}
```


# Fitness

```{r}

sum(yearFitness) # expected fitness
```

# Main growth plot in reversible and structural mass

```{r fig.width=5, fig.height=3}
param_value <- NULL
param_type  <- NULL
age         <- NULL

for (yr in 1:max_age_years) {
   param_value  <- c(param_value, Res[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   param_type <- c(param_type, rep("Reversible", 365))
   param_value  <- c(param_value, Str[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   param_type <- c(param_type, rep("Structure", 365))
   df_cost = reproCost(Str[yr,])
   param_value <- c(param_value, df_cost)
   age   <- c(age, 1:365 + 365*(yr-1))
   param_type <- c(param_type, rep("Cost", 365))
}

df_mass <- tibble(age = age, type = param_type, value = param_value)

ggplot(df_mass, aes(x = age, y = value, color = type)) + 
  xlab("Age (d)") + ylab("Mass/cost (g)") + ylim(0,NA) +
  geom_line() + theme_bw()




```

# Other model outputs 

```{r fig.width=12, fig.height=8}
val  <- NULL
mtype <- NULL
age   <- NULL

for (yr in 1:max_age_years) {
   val   <- c(val, dayIntake[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Intake (g d-1)", 365))
   val   <- c(val, dayMaintenance[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Maintenance (g d-1)", 365))
   val   <- c(val, dayNetIntake[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Net Intake (g d-1)", 365))
   val   <- c(val, dayPredatorLength[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Length (m)", 365))
   val   <- c(val, daySurvival[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Survival", 365))
   val   <- c(val, natmortality[yr, ]*365)
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Nat mortality rate (y-1)", 365))
   val   <- c(val, fishmortality[yr, ]*365)
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Fishing mortality rate (y-1)", 365))
   val   <- c(val, relGrowth[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Relative growth, (%/day)", 365))
   val   <- c(val, Str[yr, ] + Res[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Total weight (g)", 365))
   val   <- c(val, Res[yr, ]/Str[yr, ])
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("RS ratio", 365))
   val   <- c(val, 100.0*(Str[yr, ] + Res[yr, ]) /
       ((100.0*dayPredatorLength[yr, ])^3.0)) # weight = g, length = cm
   age   <- c(age, 1:365 + 365*(yr-1))
   mtype <- c(mtype, rep("Condition", 365))
}

df_rate <- tibble(age = age, val = val, mtype = mtype)

ggplot(filter(df_rate, val > 0), aes(x = age/365, y = val)) + 
  xlab("Age (years)") + ylab("Value") + ylim(0,NA) +
  scale_x_continuous(breaks = seq(from = 0, to = max_age_years, by = 2)) +
  geom_line() + facet_wrap( ~ mtype, scale = "free_y") + theme_bw()

```

#weight plot for cod
```{r}
age        <- c(  1,   2,    3,    4,    5,    6,    7,    8,     9)

#values of  mininum and maximum weight observed in empirical studies (Supplementary Table 1)
weight_min <- c( 50, 200,  500, 1000, 1500, 2000, 3000, 4000,  4000) 
weight_max <- c(200, 500, 1200, 1900, 3200, 4500, 6000, 7000, 10000)


df_ref <- tibble(Year = age, min = weight_min, max = weight_max)

df_weight <- filter(df_rate, age <= 9*365, mtype == "Total weight (g)") %>%
  select(age, weight = val)
df_weight$age <- df_weight$age/365

plot(x = df_weight$age, y = df_weight$weight, type = "n", ylim = c(0,10100),
  xlab = "Age (years)", ylab = "Weight (g)")
arrows(x0 = df_ref$Year, y0 = df_ref$min, x1 = df_ref$Year, y1 = df_ref$max,
  length=0.05, angle=90, code=3, col = "grey", lwd = 2)
lines(x = df_weight$age, y = df_weight$weight, lwd = 1)
```

#length plot for cod
```{r}
age        <- c(   1,    2,    3,    4,    5,    6,    7,    8,    9)

#values of mininum and maximum length observed in empirical studies (Supplementary Table 1)
length_min <- c(0.12, 0.25, 0.35, 0.45, 0.50, 0.60, 0.65,  0.70, 0.70)
length_max <- c(0.20, 0.35, 0.45, 0.60, 0.70, 0.75, 0.85,  0.90, 1.00)
df_ref <- tibble(Year = age, min = length_min, max = length_max)

df_length <- filter(df_rate, age <= 9*365, mtype == "Length (m)") %>%
  select(age, length = val)
df_length$age <- df_length$age/365

plot(x = df_length$age, y = df_length$length, type = "n", ylim = c(0,1),
  xlab = "Age (years)", ylab = "length (m)")
arrows(x0 = df_ref$Year, y0 = df_ref$min, x1 = df_ref$Year, y1 = df_ref$max,
  length=0.05, angle=90, code=3, col = "grey", lwd = 2)
lines(x = df_length$age, y = df_length$length, lwd = 1)
```

#condition for cod
```{r}
df_condition <- filter(df_rate, age <= 9*365, mtype == "Condition") %>%
  select(age, condition = val)
df_condition$age <- df_condition$age/365

plot(x = df_condition$age, y = df_condition$condition, type = "n", 
  ylim = c(0.7,1.4), xlab = "Age (years)", ylab = "Condition")

#Empirically observed conditions are shown with horizontal lines
abline(h = 0.8, lwd = 2, col = "grey")
abline(h = 1.4, lwd = 2, col = "grey")
lines(x = df_condition$age, y = df_condition$condition, lwd = 1)
```

# Reproduction

```{r fig.width=12, fig.height=3}
n <- max_age_years
max_y <- max(c(Res[1:n,364], Str[1:n,364]))

val  <- NULL
mtype <- NULL
age   <- NULL

val   <- c(val, yearRepCost[1:n])
age   <- c(age, 1:n)
mtype <- c(mtype, rep("Reproductive cost (g)", n))

val   <- c(val, yearSpawn[1:n])
age   <- c(age, 1:n)
mtype <- c(mtype, rep("Spawn (g)", n))

val   <- c(val, yearFitness[1:n])
age   <- c(age, 1:n)
mtype <- c(mtype, rep("Expected fitness (g)", n))

df_rate <- tibble(age = age, val = val, mtype = mtype)

ggplot(filter(df_rate, val >= 0), aes(x = age, y = val)) + 
  xlab("Age (years)") + ylab("Value") + ylim(0,NA) +
  scale_x_continuous(breaks = seq(from = 0, to = max_age_years, by = 2)) +
  geom_line() + geom_point() + facet_wrap( ~ mtype, scale = "free_y") + theme_bw()

```

# Estimation and plotting of allometric relationships such as intake, maintenance and net intake against structural mass and against total mass

```{r fig.width=12, fig.height=3}
# structural mass relations

var_value <- NULL
var_type  <- NULL
mass      <- NULL
year      <- NULL

for (yr in 1:(max_age_years-1)) {
   var_value <- c(var_value, dayNetIntake[yr, ])
   var_type  <- c(var_type,  rep("Net intake (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ])
   year      <- c(year,      rep(yr, 365))
   var_value <- c(var_value, dayIntake[yr, ])
   var_type  <- c(var_type,  rep("Intake (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ])
   year      <- c(year,      rep(yr, 365))
   var_value <- c(var_value, dayMaintenance[yr, ])
   var_type  <- c(var_type,  rep("Maintenance (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ])
   year      <- c(year,      rep(yr, 365))
}

df_rate <- tibble(Val = var_value, Type = var_type, Mass = mass, Year = year)
```

```{r}
# total mass relations (structural + reserves)

var_value <- NULL
var_type  <- NULL
mass      <- NULL
year      <- NULL

for (yr in 1:(max_age_years-1)) {
   var_value <- c(var_value, dayNetIntake[yr, ])
   var_type  <- c(var_type,  rep("Net intake (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ] + Res[yr, ])
   year      <- c(year,      rep(yr, 365))
   var_value <- c(var_value, dayIntake[yr, ])
   var_type  <- c(var_type,  rep("Intake (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ] + Res[yr, ])
   year      <- c(year,      rep(yr, 365))
   var_value <- c(var_value, dayMaintenance[yr, ])
   var_type  <- c(var_type,  rep("Maintenance (g d-1)", 365))
   mass      <- c(mass,      Str[yr, ] + Res[yr, ])
   year      <- c(year,      rep(yr, 365))
}

df_rate2 <- tibble(Val = var_value, Type = var_type, Mass = mass, Year = year)
```

```{r fig.width=12, fig.height=3}
ggplot(filter(df_rate, Val != 0 & Year <= 7), 
  aes(x = Mass, y = Val, color = factor(Year))) + 
  xlab("Structural mass (g)") + ylab("Value") + ylim(0,NA) +
  labs(color = "Age\n(years)") + scale_color_brewer(palette="Spectral") +
  geom_line(size = 1) + facet_wrap( ~ Type, scale = "free_y") + 
  theme_bw() +
  theme(panel.background = element_rect(colour = "black", fill = "#E6F5FE"))

ggplot(filter(df_rate2, Val != 0 & Year <= 7), 
  aes(x = Mass, y = Val, color = factor(Year))) + 
  xlab("Total mass (g)") + ylab("Value") + ylim(0,NA) +
  labs(color = "Age\n(years)") + scale_color_brewer(palette="Spectral") +
  geom_line(size = 1) + facet_wrap( ~ Type, scale = "free_y") + 
  theme_bw() +
  theme(panel.background = element_rect(colour = "black", fill = "#E6F5FE"))
```

